---
title: "Item Screening with twigg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Item Screening with twigg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

`twigg` implements tools for item screening in graphical log-linear 
Rasch models (GLLRMs). The package follows the screening logic described by Kreiner
and Christensen: first check consistency, then screen for local
dependence (LD), differential item functioning (DIF), score covariate
associations, and finally translate the surviving evidence into graph objects.

Please note, the purpose of this vignette is not to give a full statistical 
deriviation of the procedure but to showcase how the package works in practice.

The main workflow is:

1. Check item consistency with `M1()`, `M2()`, `M3()`, or `screen_items()`.
2. Screen for potential LD with `screen_LD()`.
3. Screen for potential DIF with `screen_DIF()`.
4. Remove spurious LD and DIF evidence with `genuine_LD()`, `run_step3b()`,
   `run_step3c()`, and `combine_step3bc()`.
5. Screen score covariate associations with `step4_structure_screen()`.
6. Build and check the GLLRM graph with `build_gllrm_graph()` and
   `step6_check_gllrm()`.

The package returns ordinary data frames at each step. This makes the results
easy to inspect, report, and most importantly; easy to pass to as arguments
to other existing (different) package functions, specifically plot functions.

```{r setup}
library(twigg)
```

## Data

The package includes some small simulated datasets for examples. Here we use
`toy_sp_DIF_and_LD`, which contains five ordinal pain items and two exogenous
covariates.

```{r data}
data(toy_sp_DIF_and_LD)

items <- paste0("pain", 1:5)
covariates <- c("sex", "age")

str(toy_sp_DIF_and_LD[, c(items, covariates)])
```

## Step 1: Consistency Checks

The first screening step checks whether the item responses behave in a way that
is compatible with a Rasch-type measurement model. The convenience function
`screen_items()` combines the package's M1, M2, and M3 checks.

```{r step1, eval = FALSE}
screen_items(
  dataset = toy_sp_DIF_and_LD,
  items = items,
  covariates = covariates
)
```

We can also run only M1 and M3 inside the `screen_items()` function:

```{r step1 niche example, eval = FALSE}
screen_items(
  dataset = toy_sp_DIF_and_LD,
  items = items,
  covariates = covariates,
  run_M1 = TRUE,
  run_M2 = FALSE,
  run_M3 = TRUE
)
```

The components can also be run directly:

```{r step1-components, eval = FALSE}
M1(toy_sp_DIF_and_LD, items)
M2(toy_sp_DIF_and_LD, items)
M3(toy_sp_DIF_and_LD, items, covariates)
```

## Step 2: Initial LD and DIF Screening

LD screening tests conditional associations between item pairs given rest
scores. DIF screening tests conditional associations between items and
exogenous varaibles given the total score.

```{r step2-ld, eval = FALSE}
ld_screen <- screen_LD(
  dataset = toy_sp_DIF_and_LD,
  items = items,
  crit_val = 0.05
)
```

```{r step2-dif, eval = FALSE}
dif_screen <- screen_DIF(
  dataset = toy_sp_DIF_and_LD,
  items = items,
  covariates = covariates,
  crit_val = 0.05
)
```

## Step 3: Remove Spurious LD and DIF Evidence

The initial LD and DIF screens can contain spurious evidence because the tests
are derived under a simpler Rasch graph. Step 3 tries to separate genuine from
spurious evidence.

For LD, `genuine_LD()` takes the output of `screen_LD()` and returns item pairs
that survive the Step 3a elimination procedure.

```{r step3-ld, eval = FALSE}
ld_step3 <- genuine_LD(ld_screen)
ld_step3$genuine_ld
```

For DIF, we first creates the `SOURCE` and `DIF` lists, then check
both item-specific source sets and source-specific DIF item sets.

```{r step3-dif, eval = FALSE}
sd <- s.d_list(
  screen_DIF_output = dif_screen,
  items = items,
  covariates = covariates
)

step3b <- run_step3b(
  dataset = toy_sp_DIF_and_LD,
  source_list = sd$SOURCE,
  items = items,
  B = 1000
)

step3c <- run_step3c(
  dataset = toy_sp_DIF_and_LD,
  dif_list = sd$DIF,
  items = items,
  B = 1000
)

dif_step3 <- combine_step3bc(
  step3b_results = step3b,
  step3c_results = step3c,
  original_source_list = sd$SOURCE,
  original_dif_list = sd$DIF
)

dif_step3$table
```

## Step 4: Score-Covariate Associations

Step 4 screens structural associations between the total score and the
exogenous covariates. It is a backwards elimination procedure: each currently
retained covariate is tested against the total score while conditioning on the
other retained covariates.

If `adjust_method` is supplied, adjusted p-values are used for the elimination
decision. Otherwise, raw p-values are used.

```{r step4, eval = FALSE}
step4 <- step4_structure_screen(
  data = toy_sp_DIF_and_LD,
  items = items,
  covariates = covariates,
  alpha = 0.05,
  adjust_method = NULL,
  B = 1000
)

step4$criterion_validity
step4$criterion_validity_comparison
```

## Step 5: Build the GLLRM Graph

Step 5 collects the retained LD, DIF, and score covariate evidence into a graph
representation. The main `edges` table is intentionally minimal: it contains
only `from`, `to`, and `edge_type`. The statistical evidence remains in
`ld_edges`, `dif_edges`, and `score_edges`.

The example below constructs a small Step 5 graph directly from data frames.
This keeps the vignette fast while showing the shape of the object returned by
`build_gllrm_graph()`.

```{r step5-example}
ld_example <- data.frame(
  item1 = "pain1",
  item2 = "pain2",
  weighted_partial_gamma = 0.35
)

dif_example <- data.frame(
  item = "pain3",
  DIF_source = "sex",
  gamma_1 = 0.22,
  p_value_1 = 0.01,
  conditioned_on_1 = "Score",
  gamma_2 = 0.20,
  p_value_2 = 0.02,
  conditioned_on_2 = "Score",
  conclusion = "DIF"
)

step4_example <- data.frame(
  covariate = c("sex", "age"),
  gamma = c(0.30, 0.25),
  p_value = c(0.01, 0.03),
  adjusted_p_value = c(NA_real_, NA_real_),
  conditioned_on = c("age", "sex"),
  supports_criterion_validity = c(TRUE, TRUE)
)

step5 <- build_gllrm_graph(
  items = items,
  covariates = covariates,
  ld = ld_example,
  dif = dif_example,
  step4 = step4_example
)

step5
step5$nodes
step5$edges
```

## IRT and Moralized Graphs

The Step 5 graph records the LD, DIF, and score-covariate edges supported by the screening procedure. It is therefore an initial evidence graph. Two helper functions 
derive the graph forms used in the GLLRM screening article.

`build_irt_graph()` creates an IRT-style chain graph:

- the latent variable points to every item;
- retained score-covariate associations become covariate-to-latent arrows;
- DIF becomes covariate-to-item arrows;
- LD remains undirected between items.

```{r irt-graph}
irt_graph <- build_irt_graph(step5, theta = "theta")

irt_graph$nodes
irt_graph$edges
```

`build_moralized_graph()` creates the moralized marginal graph used to read
off global Markov property hypotheses. It replaces the latent variable with a
total score node, adds item-score moralization edges, and makes all edges
undirected.

```{r moral-graph}
moral_graph <- build_moralized_graph(step5, score_node = "#")

moral_graph$nodes
moral_graph$edges
```

These tables can be passed to graph packages. For example, the moralized graph
can be drawn using the `simpleNetwork` function form the `networkD3` package.

```{r igraph-example, eval = FALSE}
g_moral <- networkD3::simpleNetwork(moral_graph$edges[,c("from", "to")], 
                              height="100px", 
                              width="100px", 
                              zoom = T)

g_moral
```

The IRT graph has both directed and undirected edges so it is not as stright 
forward. However the `directed` column in `irt_graph$edges` is included so 
potential custom plotting code can draw arrows only for the directed rows.

## Step 6: Check the GLLRM

Step 6 retests the retained Step 5 DIF and LD edges under conditioning sets
implied by the Step 5 graph. The current implementation provides the first 
Step 6 layer:

- C5-style checks for retained DIF edges;
- C7/C9-style checks for retained LD edges;
- removal of unsupported DIF and LD edges;
- construction of an updated final graph and moralized graph.

```{r step6, eval = FALSE}
step6 <- step6_check_gllrm(
  data = toy_sp_DIF_and_LD,
  step5 = step5,
  alpha = 0.05,
  B = 1000
)

step6$dif_tests
step6$ld_tests
step6$removed_dif
step6$removed_ld
step6$final_graph
step6$moralized_graph
```

Step 6 uses the same partial gamma conditional independence modules as the DIF,
LD, and step 4. Searching the moralized graph for all minimal separating sets 
is not yet implemented.

## Putting the Workflow Together

A complete analysis script has the following shape:

```{r full-workflow, eval = FALSE}
data(toy_sp_DIF_and_LD)

items <- paste0("pain", 1:5)
covariates <- c("sex", "age")

screen_items(toy_sp_DIF_and_LD, items, covariates)

ld_screen <- screen_LD(toy_sp_DIF_and_LD, items)
ld_step3 <- genuine_LD(ld_screen)

dif_screen <- screen_DIF(toy_sp_DIF_and_LD, items, covariates)
sd <- s.d_list(dif_screen, items, covariates)
step3b <- run_step3b(toy_sp_DIF_and_LD, sd$SOURCE, items, B = 1000)
step3c <- run_step3c(toy_sp_DIF_and_LD, sd$DIF, items, B = 1000)
dif_step3 <- combine_step3bc(step3b, step3c, sd$SOURCE, sd$DIF)

step4 <- step4_structure_screen(
  data = toy_sp_DIF_and_LD,
  items = items,
  covariates = covariates,
  B = 1000
)

step5 <- build_gllrm_graph(
  items = items,
  covariates = covariates,
  ld = ld_step3,
  dif = dif_step3,
  step4 = step4
)

irt_graph <- build_irt_graph(step5)
moral_graph <- build_moralized_graph(step5)

step6 <- step6_check_gllrm(
  data = toy_sp_DIF_and_LD,
  step5 = step5,
  B = 1000
)
```

The natural final object to inspect, as of now, is `step6$final_graph`, 
together with the test tables documenting which edges were retained or removed.
